---
output: md_document
---

<!-- Please edit README.Rmd - not README.md -->

[Presentation](https://tad-week-06-data-import-export.netlify.app)

# Week 6: Data import / export 

This week we'll discuss different methods for import/exporting a range of file types.


# Resources

- [R for Data Science](https://r4ds.had.co.nz/transform.html) has some bits on loading/exporting data
- [Data import/export cheat sheet](https://github.com/rstudio/cheatsheets/blob/main/data-import.pdf)

# Exercises

1.  **Setup/Data import and library(s)**

        library(readxl)
        library(readr)
        library(data.table)
        library(readxl)
        library(DBI)
        library(odbc)       
        library(janitor)
        library(dplyr)
        library(fst)

2.  **Content**

    Today we will aim to cover and mainly focus on:

    We will cover

    -   Importing/writing .csv and .xlsx files
    -   `sie()` function
    -   Importing/writing .fst files
    -   Connecting to SQL
    -   Downloading table/views from SQL
    -   Writing to tables to SQL
    -   Import files from other software packages.
    
3.  **Import .csv and .xlsx**

    3.1 Import a CSV file into R.

    3.2 Import a XLSX file into R.

    3.3 Write a CSV.

    3.4 Use the `sie()` function to view a data frame in Excel without writting one as a CSV or XSLX. Add the below code to your R profile.  
    
    ``` {r eval = FALSE}
    
    shortcut_env <- new.env()
    with(.shortcut_env, {
      shortcut <- function(f) structure(f, class = "shortcut")
      print.shortcut <- function(f, ...) {
        res <- withVisible(f(...))
        if (res$visible) {
          print(res$value)
        }
        res$value
      }
  
  
      sie <- shortcut(function(.data){
        tmp <- paste0(tempfile(), ".xlsx")
        writexl::write_xlsx(.data, tmp)
        shell.exec(tmp)
      })
  
    })



    library(stats)
    suppressMessages(attach(.shortcut_env))
    registerS3method("print", "shortcut", print.shortcut)
    
    ```


4.  **Connect to an SQL database**

    4.1 Conncet to an SQL database that you have access to. **All TAD staff should have access to [TAD_UserSpace]**.
    
    4.2 You will need to create a config.yml file. Look at the presentation to help you. See example below:
    
    In config.yml
    
    ``` {r eval = FALSE}
    default:
      tad_user_sql:
        driver: "SQL Server Native Client 11.0"
        server: "3DCPRI-PDB16\\ACSQLS"
        database: "TAD_UserSpace"
        uid: ""
        pwd: ""
        trusted: "Yes"
        encoding: "LATIN1"`
        
    ```
        
    Use connection function then line below
    
    tad_user_con <- connection_function(tad_user_sql, "TAD_UserSpace")


5.  **Download from data from SQL connection**

    5.1 Download data from SQL with the conncetion you just made. You can do this any other the ways explained in presenation.


6.  **`.fst` and `haven`**

    6.1 Create a .fst file from a large file you have. See how much quicker it is to load.
    6.2 If you have an SPSS, STATA or SAS file try and use `haven` package to read in the data.
